Hydra Battles and Cie

### 2021-08-09 doc/chapter-powers.tex
fix typos in chapter-powers.tex and chapter-primrec.tex

In~\cite{BC04}, several exercices~\footnote{Exercises 9.8 (page 270), 9.10 (page 271), 9.15 (page 276), 9.17 (page 284), and 15.8 (page 418).}  
present ways to compute Fibonacci numbers, with the less number of recursive calls  as possible. Please note that these optimizations and the formal proof of their correctness are \emph{ad-hoc}, \emph{i.e.} exclusively written for the 
Fibonacci numbers.
In contrast, the optimizations we present in this document apply, in their vast majority, \emph{generic} techniques of efficient computation of powers in a monoid. 
This example of Fibonacci numbers has been developed with Yves Bertot, who wrote a first version with \texttt{SSreflect/Mathcomp}~\cite{MCB}.

@@

Thus, any function able to compute more or less efficiently powers in a monoid will
give an algorithm for computing Fibonacci numbers. Unlike the \emph{ad-hoc} aforementionned proofs of~\cite{BC04}, the correctness of such an algorithm is a direct consequence 
of the correctness of the used powering function.
Several examples will be presented in the rest of this document
(in sections~\vref{sect:fibonacci-pos-bpow}).

@@

  In the following inductive type definition, the intended meaning 
  of the contruct (\texttt{Mult $x$ $y$ $k$})  is \emph{``multiply \texttt{x} with 
\texttt{y}, then send  the result of this multiplication to 
  the continuation  \texttt{k}''}.

@@

The following \emph{monadic} 
notation makes terms of type \texttt{computation} look like
expressions of a small programming language dedicated to sequences of mutiplications.
Please look at \emph{CPDT}~\cite{chlipalacpdt2011} for more details on monadic notations in \coq.
\label{monadic-mult}

@@

\subsection{Correctness proofs by reflection}
\label{reflection-section}
\index{coq}{Proofs by reflection}
Instead of letting the tactic \texttt{rewrite} look for contexts in which
setoid rewriting is possible, we propose to use (deterministic) computations for
obtaining a ``canonical'' form for terms generated from a variable \texttt{x}
by contructors associated with monoid multiplication and neutral element.

@@

\subsubsection{Definition}
A chain $c$ is \emph{parametric} if it has the same behaviour for any
pair of types $A$  and $B$, any relation $R$
between  $A$ and $B$ and any $R$-related pair of 
arguments $a$ and $b$:

@@

Unfortunately, our construct is still very inefficient, since it results in 
duplications of computations, as shown by the normal form of \texttt{F23}.

@@

We are now ready to define various operators on f- and k-chains, and prove these
operators preserve correcness and properness. We will also show that 
these operators allow to generate easily correct chains for any positive 
exponent. They will be used to generate chains for
numbers of the form $n=bq+r$ where $0\leq r < b$, assuming the previous
construction of correct chains for $r$, $b$ and $q$.
For instance, Figure~\ref{fig:K7-3-decomposition} shows how \texttt{K7\_3} is built
as a composition of \texttt{K3\_1} and \texttt{F2}.

@@

Since our construction is based on Euclidean division, we could not
define our chain generator by structural recursion. 
For simplicity's sake, we chosed to avoid dependent elimination
 and used \texttt{Function}  with a decreasing measure.

@@

We had two slight technical problems to consider:
\begin{itemize}
\item The generation of a k-chain for $n$ and $p$ is meaningfull only if $p < n$. Thus, in order to avoid a clumsy  dependent pattern-matching, we chosed to represent 
     a pair $(n,p)$ where $0<p<n$ by a pair of positive numbers $(p,d)$ where 
     $d=n-p$
\item In order to avoid to deal explicitly with mutual recursion, we
     defined a type called \texttt{signature} for representing both
     forms of function calls.
     Thus, it is easy to define a decreasing measure on type 
     \texttt{signature} for proving termination. 
    Likewise, correctness and properness statements are also indexed by 
    this type.

@@

\subsection{Generation of chains using Euclidean Dibision}

@@

\begin{enumerate}
\item  Prove that the chain generated by \texttt{Fexp2} is optimal.
\item Prove that  the length of any optimal chain for $n$ is
greater than or equal to $\floor{\log_2{n}}$.
\item Prove that, for any positive $n$, the length of any Euclidean chain generated by the 
  dichotomic strategy  is always less than or equal to
  the length of \texttt{binary\_chain $n$}, and for an infinite number
of positive integers $n$, the first chain  is strictly shorter
than  the latter.
\item Prove that our implementation of the dichotomic strategy describes
 the same function as in the litterature (for instance ~\cite{DBLP:journals/ita/BrlekCHM95}.)
This is important if we want to follow the complexity analyses in this and similar articles.
\item Study how to \emph{compile} a chain into imperative code, using a register allocation strategy (it may be useful  to define \emph{chain width} ).

>>>>

In~\cite{BC04}, several exercises~\footnote{Exercises 9.8 (page 270), 9.10 (page 271), 9.15 (page 276), 9.17 (page 284), and 15.8 (page 418).}
present ways to compute Fibonacci numbers, with the less number of recursive calls  as possible. Please note that these optimizations and the formal proof of their correctness are \emph{ad-hoc}, \emph{i.e.} exclusively written for the 
Fibonacci numbers.
In contrast, the optimizations we present in this document apply, in their vast majority, \emph{generic} techniques of efficient computation of powers in a monoid. 
This example of Fibonacci numbers has been developed with Yves Bertot, who wrote a first version with \texttt{SSreflect/Mathcomp}~\cite{MCB}.

@@

Thus, any function able to compute more or less efficiently powers in a monoid will
give an algorithm for computing Fibonacci numbers. Unlike the \emph{ad-hoc} aforementioned proofs of~\cite{BC04}, the correctness of such an algorithm is a direct consequence
of the correctness of the used powering function.
Several examples will be presented in the rest of this document
(in sections~\vref{sect:fibonacci-pos-bpow}).

@@

  In the following inductive type definition, the intended meaning 
  of the construct (\texttt{Mult $x$ $y$ $k$})  is \emph{``multiply \texttt{x} with
\texttt{y}, then send  the result of this multiplication to 
  the continuation  \texttt{k}''}.

@@

The following \emph{monadic} 
notation makes terms of type \texttt{computation} look like
expressions of a small programming language dedicated to sequences of multiplications.
Please look at \emph{CPDT}~\cite{chlipalacpdt2011} for more details on monadic notations in \coq.
\label{monadic-mult}

@@

\subsection{Correctness proofs by reflection}
\label{reflection-section}
\index{coq}{Proofs by reflection}
Instead of letting the tactic \texttt{rewrite} look for contexts in which
setoid rewriting is possible, we propose to use (deterministic) computations for
obtaining a ``canonical'' form for terms generated from a variable \texttt{x}
by constructors associated with monoid multiplication and neutral element.

@@

\subsubsection{Definition}
A chain $c$ is \emph{parametric} if it has the same behavior for any
pair of types $A$  and $B$, any relation $R$
between  $A$ and $B$ and any $R$-related pair of 
arguments $a$ and $b$:

@@

Unfortunately, our construct is still very inefficient, since it results in 
duplication of computations, as shown by the normal form of \texttt{F23}.

@@

We are now ready to define various operators on f- and k-chains, and prove these
operators preserve correctness and properness. We will also show that
these operators allow to generate easily correct chains for any positive 
exponent. They will be used to generate chains for
numbers of the form $n=bq+r$ where $0\leq r < b$, assuming the previous
construction of correct chains for $r$, $b$ and $q$.
For instance, Figure~\ref{fig:K7-3-decomposition} shows how \texttt{K7\_3} is built
as a composition of \texttt{K3\_1} and \texttt{F2}.

@@

Since our construction is based on Euclidean division, we could not
define our chain generator by structural recursion. 
For simplicity's sake, we chose to avoid dependent elimination
 and used \texttt{Function}  with a decreasing measure.

@@

We had two slight technical problems to consider:
\begin{itemize}
\item The generation of a k-chain for $n$ and $p$ is meaningful only if $p < n$. Thus, in order to avoid a clumsy  dependent pattern-matching, we chose to represent
     a pair $(n,p)$ where $0<p<n$ by a pair of positive numbers $(p,d)$ where 
     $d=n-p$
\item In order to avoid to deal explicitly with mutual recursion, we
     defined a type called \texttt{signature} for representing both
     forms of function calls.
     Thus, it is easy to define a decreasing measure on type 
     \texttt{signature} for proving termination. 
    Likewise, correctness and properness statements are also indexed by 
    this type.

@@

\subsection{Generation of chains using Euclidean Division}

@@

\begin{enumerate}
\item  Prove that the chain generated by \texttt{Fexp2} is optimal.
\item Prove that  the length of any optimal chain for $n$ is
greater than or equal to $\floor{\log_2{n}}$.
\item Prove that, for any positive $n$, the length of any Euclidean chain generated by the 
  dichotomic strategy  is always less than or equal to
  the length of \texttt{binary\_chain $n$}, and for an infinite number
of positive integers $n$, the first chain  is strictly shorter
than  the latter.
\item Prove that our implementation of the dichotomic strategy describes
 the same function as in the literature (for instance ~\cite{DBLP:journals/ita/BrlekCHM95}.)
This is important if we want to follow the complexity analyses in this and similar articles.
\item Study how to \emph{compile} a chain into imperative code, using a register allocation strategy (it may be useful  to define \emph{chain width} ).
